Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-15 programlogic.tex
various space tweaks...

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another variety of pointwise lifting that already exists in the main \iris
  distribution~\cite{dang2019rustbelt,dang2022compass}. This is essentially true, but the standard formalization does not work for index types which
  are indexed in the sense of \coq's metatheory, as our \lstinline|word n| types are.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification, we define our logic using Hoare doubles:%\footnote{This
% omits some low-level Iris details (stuckness, observations) that play no meaningful
% role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is Iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} an explicit parameter to the double, rather than simply using register assertions
directly 
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }

>>>>

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another pointwise lifting already in  \iris~\cite{dang2019rustbelt,dang2022compass}. 
  This similarity is real, but the existing lifting does not work with indexed \coq types
  like our \lstinline|word n| as a domain.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

\subsection{Selected Logical Rules}
\label{sec:selected_rules}
% Per the discussion in Section \ref{sec:issues}, w
As common for assembly-level verification, we define our logic using Hoare doubles:%\footnote{This
% omits some low-level Iris details (stuckness, observations) that play no meaningful
% role in our development.}
  \[
  \begin{array}{l}
    %\textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rtv } : \textsf{iProp }\Sigma := \\
    \{ \Phi \}_\mathsf{ rtv }\;\textsf{e} : \textsf{iProp }\Sigma := 
   % \qquad
   ((\textsf{cr3} \mapsto_{\textsf{r}} \textsf{rtv} \ast \Phi) \textsf{ rtv}) \wand \textsf{WP e } \{\_, \textsf{True} \}
    \end{array}
  \]
Our Hoare doubles $\{\Phi\}_\textsf{rtv}\;\textsf{e}$ state that the expression (i.e., sequence of instructions)
\textsf{e} are safe to execute (will not fault)
when executed with \textsf{vProp} precondition $\Phi\ast\textsf{cr3}\mapsto_{\textsf{r}} \textsf{rtv}$.
\textsf{WP} is Iris's own weakest precondition modality, unmodified~\cite{jung2018iris}.
Making \textsf{rtv} a parameter to the double (vs. a simple register assertion)
makes it possible to ensure ownership of the \lstinline|cr3| register and its value is accounted for
while avoiding some technical headaches with trying to enforce that $\Phi$ itself contains that.
\looseness=-1
% solves a technical problem with ensuring that the page table root used to evaluate
% the \textsf{vProp} (i.e., evaluating the assertion in the \emph{current}) address space
% is feasible.
% \footnote{Consider the difficulty of selecting the correct page table root value from an arbitrary
% opaque $\Phi$, which may even existentially quantify the page table root. An alternative is to
% require $\Phi$ to have a syntactic form where we can directly extract the value of \lstinline|cr3|,
% but this makes using Iris Proof Mode (IPM)~\cite{Krebbers:2017:IPH:3009837.3009855} with \textsf{vProps}
%   difficult; IPM works for any type matching the signature of an Iris \lstinline|bi|, which includes
%   \textsf{vProp}s, but manually guiding IPM to put an assertion in a specific position over and over adds
%   significant proof burden.
% }
